---
title: "Intro to Epigenomics"
author: Wade Boohar
date: 11/03/24
updated: 03/07/24
---


```{r setup}
 knitr::opts_knit$set(root.dir = normalizePath("/home1/sskatta/490_cluster/analysis_data"))
```

Package Download and Data-cleaning
```{r}
if (!require("sesameData", quietly = TRUE))
BiocManager::install("sesameData")

if (!require("sesame", quietly = TRUE))
BiocManager::install("sesame")

if (!require("limma", quietly = TRUE))
BiocManager::install("limma")
```


Load in all necessary packages
```{r}
library(TCGAbiolinks)
library(sesame)
library(sesameData)
library(limma)
```

```{r}
query <- GDCquery(project = "TCGA-BRCA",
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450",
                  data.format = "Dataframe")
GDCdownload(query)
methylation450 <- GDCprepare(query)
```

```{r}
methylation_clinical <- as.data.frame(methylation450@colData)
betas <- as.data.frame(methylation450@assays@data@listData)
cpg_sites <- as.data.frame(methylation450@rowRanges@elementMetadata)

column_mask <- ifelse(colnames(methylation_clinical) %in% c('treatments', 'primary_site', 'disease_type','sites_of_involvement'), F, T)

methylation_clinical <- methylation_clinical[,column_mask]
write.csv(methylation_clinical, 'brca_methylation_clinical.csv')


site_mask <- !grepl('-', cpg_sites$gene) & !grepl(';', cpg_sites$gene) & !is.na(cpg_sites$gene) & complete.cases(betas)
betas <- betas[site_mask,]
cpg_sites <- cpg_sites[site_mask,]

write.csv(betas, 'brca_methylation_betas.csv')
write.csv(cpg_sites, 'brca_cpg_sites.csv')
```

```{r}
methylation_clinical <- read.csv('brca_methylation_clinical.csv', row.names = 1)
betas <- read.csv('brca_methylation_betas.csv', row.names = 1)
cpg_sites <- read.csv('brca_cpg_sites.csv', row.names = 1)
```

```{r}
methylation_clinical <- read.csv("/project/rohs_1070/analysis_data/brca_methylation_clinical.csv", row.names = 1)

cpg_sites <- read.csv("/project/rohs_1070/analysis_data/brca_cpg_sites.csv", row.names = 1)


install.packages("data.table")
library(data.table)

betas <- fread('/project/rohs_1070/analysis_data/brca_methylation_betas.csv', showProgress = TRUE)

betas <- as.data.frame(betas)
rownames(betas) <- betas[[1]]
betas[[1]] <- NULL
```



(1) Naive Differential Methylation
```{r}

na_mask <- !is.na(methylation_clinical$definition)
methylation_clinical <- methylation_clinical[na_mask,]
betas_clean <- betas[,na_mask]


metastatic_mask <- methylation_clinical$definition == "Metastatic"
methylation_clinical$metastatic_binary <- !metastatic_mask


mval <- t(apply(betas_clean, 1, function(x) log2(x/(1-x)))) 

design <- model.matrix(~metastatic_binary, data = methylation_clinical)
fit <- lmFit(mval, design)
fit2 <- eBayes(fit)
```

```{r}
#Extracting model into dataframe
library(ggplot2)
dat <- data.frame(foldchange = fit[["coefficients"]][,2], logPvalue =  -log10(p.adjust(fit2[["p.value"]][,2],method='BY')), geneName = cpg_sites$gene)
dat$threshold <- as.factor(abs(dat$foldchange) < 1)

#Visualization
cols <- c("TRUE" = "grey", "FALSE" = "blue")
ggplot(data=dat, aes(x=foldchange, y = logPvalue, color=threshold)) +
  geom_point(alpha=.2, size=0.6) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour="#990000", linetype="dashed") + 
  geom_vline(xintercept = - 1, colour="#990000", linetype="dashed") +
  geom_hline(yintercept = 2, colour = "#990000", linetype="dashed") +
  theme(legend.position="none") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  theme_bw() +
  theme(legend.position = "none")
```


From the graph produced, there are significant differences in methylation patterns between metastatic and non-metastatic samples. Methylation alterations often contribute to gene silencing or activation that can influence cancer progression, and hypermethylation can be associated with metastasis. Conversely, hypomethylation of oncogenes can be observed in metastatic tumors. In terms of the methylation patterns, oncogenes and pro-metastatic genes may be upregulated due to hypomethylation. For downregulation due to hypermethylation, this may be seen through tumor-suppressor and apoptosis-related genes. This difference of methylation may be observed across the metastatic and non-metastatic samples in the volcano plot, with upregulated genes on the right and downregulated ones on the left. 


(2) Direct comparison of methylation status to transcriptional activity

#...INSERT DESeq2 Stuff here to generate 'results'...
```{r}
library(maftools)
library(DESeq2)


#rna_clinical <- rna_se@colData

rna_clinical <- read.csv("brca_rna_clinical_data.csv", row.names = 1, check.names = FALSE)
rna_genes <- read.csv("brca_rna_gene_data.csv")
rna_counts <- read.csv("brca_rna_count_data.csv", row.names = 1, check.names = FALSE)

rna_clinical$definition <- factor(rna_clinical$definition)
rna_clinical <- rna_clinical[!is.na(rna_clinical$definition), ]

rna_counts <- rna_counts[, rownames(rna_clinical)]

rna_counts_numeric <- rna_counts[, -1]
rna_counts_numeric <- as.data.frame(lapply(rna_counts_numeric, as.numeric))
keep_genes <- rowSums(rna_counts_numeric) >= 20
filtered_rna_counts <- rna_counts[keep_genes, ]

dds <- DESeqDataSetFromMatrix(countData = filtered_rna_counts,
                              colData = rna_clinical,
                              design= ~definition)

dds <- dds[rowSums(counts(dds)) >= 20, ]

dds <- DESeq(dds)

resultsNames(dds) 

results <- results(dds, format = "DataFrame", contrast = c("definition", "Primary solid Tumor", "Metastatic"))

results <- data.frame(results)

results <- results[!is.na(results$padj) & results$padj < 0.05 & abs(results$log2FoldChange) >= 1, ]

volcano_data <- data.frame(
  log2FoldChange = results$log2FoldChange,
  logPvalue = -log10(results$padj), 
  gene = rownames(results)
)

results$`-log10(padj)` <- -log10(results$padj)
results$gene_name <- rna_genes$gene_name[match(rownames(results), rna_genes$gene_id)]
significant <- sum(results$padj < 0.05, na.rm = TRUE)

EnhancedVolcano(results,
                lab = results$gene_name,
                x = 'log2FoldChange',
                y = '-log10(padj)',
                title = 'Methylation Status to Transcriptional Activity (Metastatic vs Non Metastatic)',
                pointSize = 1.0,
                labSize = 5.0)

print(volcano_plot)

```


This uses gene expression data to examine the correlation between methylation status and transcriptional activity. As mentioned earlier, hypermethylation of promoter regions in metastatic patients may silence tumor suppressor genes, leading to lower transcriptional activity of these genes. On the other hand, hypomethylation in the regulatory regions of oncogenes might lead to increased expression, facilitating tumor progression and metastasis. These may correlate to the upregulation and downregulation described in part 1. For metastatic samples, methylation may further silence tumor suppressors and potentially activating oncogenes in metastatic cases. 



```{r}
#you can also try looking at "upregulated" or "hypermethylated" !
downregulated <- results[(results$log2FoldChange < 3), 'gene_name']
hypomethylated <- dat[dat$foldchange < -1, 'geneName']
interest_genes <- intersect(downregulated, hypomethylated)
```


(Extra) Making Boxplots
```{r}
GENE<-"SCTR"

gene_counts_mask <- rna_genes$gene_name == GENE
gene_betas_mask <- cpg_sites$gene == GENE

rna_clinical_tumor <- rna_clinical$definition == "Primary solid Tumor"
methylation_clinical_tumor <- methylation_clinical$definition == "Primary solid Tumor"

rna_clinical_normal <- rna_clinical$definition == "Solid Tissue Normal"
methylation_clinical_normal <- methylation_clinical$definition == "Solid Tissue Normal"

rna_tumor <- as.numeric(rna_counts[gene_counts_mask, rna_clinical_tumor])
methylation_tumor <- (betas[gene_betas_mask, methylation_clinical_tumor])

rna_normal <- as.numeric(rna_counts[gene_counts_mask, rna_clinical_normal])
methylation_normal <- (betas[gene_betas_mask, methylation_clinical_normal])
```

```{r}
boxplot(rna_normal, rna_tumor, xlab='Group', ylab='Counts', names=c('Normal', 'Tumor'))
```


3)

TSPAN6: This gene acts as a tumor suppressor in certain cancers. Due to this, we can assume that it is downregulated in metastatic samples due to hypermethylation. The resaerch article of https://www.nature.com/articles/s41388-022-02223-y talks about how it is a suppressor of Ras-driven cancer. Given what I observed and know about hypermethylation during cancer, this would confirm my conclusion that it is downregulated. However, another article mentions in https://pmc.ncbi.nlm.nih.gov/articles/PMC11520469/ that it was actually upregulated for certain other cancers involving glioma tissues. Thus, this differs depending on the specific cancer in question.


BRCA1:
This gene is a tumor suppressor, and it would serve as such for these samples. Due to this, it would be hypermethylated in metastatic samples. Thus, we can conclude that there would be downregulation as a result.


MLH1:
This gene serves as a DNA repair gene. Once again, due to this, it would be hypermethylated in the metastatic samples that we observe. As a result, there would be downregulation of it.

